---
title: "Estimating FRTC Biomass and Carbon from a Forest Inventory"
author: "Santosh Ayer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating FRTC Biomass and Carbon from a Forest Inventory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nepalallometry)
```

## Purpose

This vignette demonstrates the complete `nepalallometry` workflow: preparing
tree inventory data, estimating FRTC-supported biomass and carbon, summarizing
results, handling unsupported species, and exporting one Excel workbook.

The implemented response is total aboveground biomass **above 0.30 m**. The
0-0.30 m stump is excluded.

## 1. Check supported species

```{r species}
frtc_species()
```

Use the standardized lowercase identifier (for example,
`schima_wallichii`) or the supported Nepali name (for example,
`chilaune`). An unsupported species remains in the output but receives
`NA` biomass.

## 2. Prepare inventory data

Each row represents one tree. The following columns are required:

- `tree_id`: unique tree identifier;
- `plot_id`: plot identifier;
- `plot_area_ha`: sampled plot area in hectares;
- `species`: standardized scientific identifier or supported Nepali name;
- `dbh_cm`: diameter at breast height in centimetres;
- `height_m`: total tree height in metres.

```{r inventory}
inventory <- data.frame(
  tree_id = 1:4,
  plot_id = c("p01", "p01", "p02", "p02"),
  plot_area_ha = c(0.05, 0.05, 0.10, 0.10),
  species = c("sal", "chilaune", "pinus_roxburghii", "acacia_catechu"),
  dbh_cm = c(30, 25, 30, 20),
  height_m = c(20, 18, 20, 15)
)

inventory
```

Plot area must be positive and constant within each plot. Tree identifiers
must be present and unique.

## 3. Estimate tree biomass and carbon

```{r tree-estimates}
tree_results <- estimate_frtc_biomass(inventory)
tree_results
```

The result includes standardized species names, FRTC density provenance,
biomass in kg/tree, carbon in kg/tree, basal area, model source, and estimation
status. Carbon is calculated using a constant fraction of 0.47.

The fourth tree is unsupported. A warning is expected, its biomass is `NA`,
and its status is `unsupported_species`. The package does not substitute a
model from another species.

For supported trees, inspect `within_calibration_range` and
`calibration_status`. The package compares each DBH and height with the
observed species-specific limits in the FRTC model-development dataset.
A prediction outside either limit is returned but identified as an
extrapolation.

```{r calibration-registry}
frtc_models()[, c(
  "species_id", "sample_size",
  "dbh_min_cm", "dbh_max_cm",
  "height_min_m", "height_max_m"
)]
```

## 4. Produce plot and forest summaries

```{r summaries}
plot_results <- frtc_plot_summary(tree_results)
species_results <- frtc_species_summary(tree_results)
dbh_results <- frtc_dbh_summary(tree_results)
forest_results <- frtc_forest_summary(plot_results)

plot_results
forest_results
```

Biomass is expressed in Mg/ha, which is numerically equal to metric tonnes per
hectare. Carbon is expressed in Mg C/ha.

Plot `p01` is a complete FRTC estimate because both trees are supported.
Plot `p02` is a partial estimate because one tree is unsupported. Its biomass
and carbon totals represent only the supported portion of that plot.

Before interpretation, inspect:

- `summary_status`;
- `unsupported_or_missing_trees`;
- `stem_coverage_pct`;
- `basal_area_coverage_pct`.

A partial estimate must not be described as whole-plot biomass.

## 5. Work from a CSV file

A blank CSV template can be created with:

```{r template, eval=FALSE}
frtc_inventory_template("frtc_inventory.csv", rows = 100)
```

After filling the template, run:

```{r csv-workflow, eval=FALSE}
results <- frtc_biomass_from_csv(
  input = "frtc_inventory.csv",
  output_format = "excel"
)

results$tree_results
results$plot_summary
results$species_summary
results$dbh_class_summary
results$forest_summary
results$files
```

The Excel workbook contains an interpretation sheet, all result tables, the
model registry, and the density rules. Use `output_format = "csv"` for
separate CSV files or `output_format = "both"` for both formats.

## 6. Inspect model information

```{r transparency}
frtc_equation("sal")
frtc_density("sal", dbh = 30)
```

Use `frtc_models()` and `frtc_density()` to inspect all equations and
recommended densities.

## Interpretation limits

The current package does not:

- estimate biomass for species outside the seven FRTC models;
- predict missing tree heights;
- estimate the excluded stump, roots, dead wood, litter, or soil carbon;
- provide formal prediction intervals or uncertainty propagation;
- convert a partial supported-species estimate into total plot biomass.

Calibration flags describe whether a prediction is inside the observed DBH and
height domain. They do not quantify prediction uncertainty or guarantee that
an in-range prediction is accurate.

Users should cite both `nepalallometry` and the original FRTC (2025) report.
