---
title: "Methods & Equations"
author: "Santosh Ayer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methods & Equations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nepalallometry)
```

This page documents the equations, coefficients, density rules, units, and calculation boundaries implemented in *nepalallometry*. The package operationalizes published methods; coefficients are not re-fitted by the package. Users should consult the original sources when a method is used for scientific or regulatory work.

## 1. Biomass and carbon

### 1.1 FRTC 2025 total aboveground biomass

FRTC models use one of two forms:

$$
B=a(D^2H\rho)^b
$$

or

$$
B=aD^bH^c
$$

where $B$ is biomass (kg tree$^{-1}$), $D$ is DBH (cm), $H$ is total height (m), $\rho$ is basic wood density (g cm$^{-3}$), and $a$, $b$, and $c$ are published coefficients.

The implemented response is total aboveground biomass **above 0.30 m**; the 0–0.30 m stump is excluded.

#### Species, equations and coefficients

The table below is generated directly from the package model registry so that the website reflects the coefficients actually used by *nepalallometry*.

```{r frtc-biomass-table, echo=FALSE}
x <- frtc_models()
x <- x[, c("scientific_name", "nepali_name", "equation", "a", "b", "c",
           "sample_size", "dbh_min_cm", "dbh_max_cm", "height_min_m",
           "height_max_m", "operational_density_basis")]
names(x) <- c("Species", "Nepali name", "Equation", "a", "b", "c", "n",
              "DBH min (cm)", "DBH max (cm)", "Height min (m)",
              "Height max (m)", "Density basis")
knitr::kable(x, digits = 6)
```

#### FRTC wood-density rules

Wood density is required only for FRTC models using $B=a(D^2H\rho)^b$. For *Castanopsis* spp. and *Pinus roxburghii*, density is not required by the selected FRTC biomass equation. *Shorea robusta* and *Terminalia alata* use DBH-class-specific densities; *Alnus nepalensis*, *Lagerstroemia parviflora*, and *Schima wallichii* use species-average densities.

```{r frtc-density-table, echo=FALSE}
d <- frtc_density()
names(d) <- c("Species ID", "DBH class (cm)", "Density (g/cm3)",
              "Density basis", "Source")
knitr::kable(d, digits = 4)
```

These density rules are from **FRTC (2025), Table 11** and are part of the FRTC biomass pathway.

### 1.2 Sharma & Pukkala 1990 biomass

The Sharma & Pukkala pathway first calculates stem volume using:

$$
V_s=\frac{\exp(a+b\ln D+c\ln H)}{1000}
$$

The original equation produces volume in dm$^3$; *nepalallometry* therefore **divides by 1000 to return m$^3$**.

Stem biomass is then:

$$
B_s=V_s\rho
$$

where $\rho$ is the air-dry wood density in kg m$^{-3}$ reported by Sharma & Pukkala (1990).

Branch and foliage biomass ratios are **species-specific and DBH-dependent**. Each species or species group has separate published small-, medium-, and large-tree ratios for branches and foliage. For species or species group $j$, *nepalallometry* first selects that group's published parameters and then calculates the applicable branch and foliage ratios independently from DBH:

$$
B_b=B_sR_{b,j}(D)
$$

$$
B_f=B_sR_{f,j}(D)
$$

and total aboveground biomass is:

$$
B_t=B_s+B_b+B_f
$$

For each component (branch or foliage), the selected species-specific ratio is interpolated according to DBH as:

$$
R_j(D)=s_j, \quad D<10
$$

$$
R_j(D)=\frac{(D-10)m_j+(40-D)s_j}{30}, \quad 10\le D\le40
$$

$$
R_j(D)=\frac{(D-40)l_j+(70-D)m_j}{30}, \quad 40<D\le70
$$

$$
R_j(D)=l_j, \quad D>70
$$

where $s_j$, $m_j$, and $l_j$ are the published small-, medium-, and large-tree ratios for species or species group $j$. **Separate parameter sets are used for branches and foliage.** Thus, the package does not apply one common branch or foliage ratio across species; it selects the species-specific parameters first and then applies the DBH interpolation.

#### Species, coefficients, densities and component ratios

```{r sp-biomass-table, echo=FALSE}
s <- sharma_pukkala_species()
s <- s[, c("scientific_name", "nepali_name", "a", "b", "c",
           "density_kg_m3", "branch_s", "branch_m", "branch_l",
           "foliage_s", "foliage_m", "foliage_l", "sample_size",
           "dbh_min_cm", "dbh_max_cm")]
names(s) <- c("Species/group", "Nepali name", "a", "b", "c",
              "Air-dry density (kg/m3)", "Branch s", "Branch m", "Branch l",
              "Foliage s", "Foliage m", "Foliage l", "n",
              "DBH min (cm)", "DBH max (cm)")
knitr::kable(s, digits = 4)
```

The table includes the **Miscellaneous species in Terai** and **Miscellaneous species in Hills** models. When a tree is not represented by a named Sharma & Pukkala model and a miscellaneous equation is appropriate, the user must explicitly identify the applicable Terai or Hills group; the package does not infer this geographic classification.

### 1.3 Chave et al. 2014

The height-inclusive pantropical equation implemented by the package is:

$$
AGB=0.0673(\rho D^2H)^{0.976}
$$

where $AGB$ is oven-dry aboveground biomass (kg tree$^{-1}$), $D$ is DBH (cm), $H$ is total height (m), and $\rho$ is basic wood density (g cm$^{-3}$), defined as oven-dry mass divided by fresh volume.

#### Where does the Chave wood density come from?

Because wood density enters the Chave equation directly, *nepalallometry* records both the density value and its provenance. The implemented lookup hierarchy is:

1. **User-supplied density.** If a non-missing `wood_density` is supplied, that value overrides the automatic lookup for that tree.
2. **FRTC 2025 density.** For the five FRTC species for which FRTC provides a recommended density for biomass estimation (*Alnus nepalensis*, *Lagerstroemia parviflora*, *Shorea robusta*, *Schima wallichii*, and *Terminalia alata*), the package uses the FRTC (2025) recommended basic density. For *Shorea robusta* and *Terminalia alata*, this is DBH-class specific; for the other three it is the FRTC species average.
3. **Global Wood Density Database v2.2 (GWDD v2.2).** If an FRTC density is not available, the package searches its bundled GWDD v2.2 tables. Matching proceeds from an accepted exact infraspecific taxon, where applicable, to binomial and then genus level. The package uses the GWDD model-derived trunk-density field `wsg_est_trunk`.
4. **No defensible match.** If no density is available after these steps, Chave biomass is returned as `NA`; the package does not invent a density.

Thus, **Chave does not automatically use one universal density source for every species**. The output records `wood_density_g_cm3`, `density_source`, `density_match_level`, `density_taxon_matched`, and `density_value_field` so the value used for each tree can be audited.

For example:

```{r chave-density-example, eval=FALSE}
chave_biomass(30, 20, "sal", keep_inputs = TRUE)
chave_biomass(30, 20, "Dalbergia sissoo", keep_inputs = TRUE)
```

The FRTC density rules shown above are therefore also the densities used first by the Chave pathway for those five FRTC species. Other taxa can use GWDD v2.2 according to the hierarchy above.

### 1.4 Carbon conversion

Carbon is derived from biomass rather than estimated by a separate allometric equation:

$$
C=AGB\times CF
$$

where $CF$ is the carbon fraction. The default in *nepalallometry* is 0.47, following IPCC (2006).

## 2. Tree volume

### 2.1 FRTC 2025 stem volume

FRTC selected equations use one of three forms:

$$
V=aD^bH^c
$$

$$
V=a(D^2H)^b
$$

or

$$
V=a+b(D^2H)
$$

where $V$ is m$^3$ tree$^{-1}$, $D$ is DBH (cm), and $H$ is total height (m).

FRTC reports three **different volume definitions**: total stem volume over bark (`total_ob`), under-bark stem volume to a 20-cm over-bark top diameter (`ub_20cm`), and under-bark stem volume to a 10-cm over-bark top diameter (`ub_10cm`). All exclude the 0.30-m stump and branches. These outputs should not be summed or averaged.

#### Species-specific FRTC volume coefficients

The following coefficients reproduce the FRTC (2025) selected equations implemented in the package.

```{r frtc-volume-table, echo=FALSE}
fv <- data.frame(
  Species = rep(c("Alnus nepalensis", "Castanopsis spp.", "Lagerstroemia parviflora",
                  "Pinus roxburghii", "Shorea robusta", "Schima wallichii",
                  "Terminalia alata"), 3),
  `Volume definition` = rep(c("Total over-bark stem", "Under-bark to 20-cm top",
                              "Under-bark to 10-cm top"), each = 7),
  Form = c(
    "a*D^b*H^c", "a*(D^2*H)^b", "a*(D^2*H)^b", "a*(D^2*H)^b", "a*(D^2*H)^b", "a*D^b*H^c", "a*(D^2*H)^b",
    "a*(D^2*H)^b", "a*(D^2*H)^b", "a+b*(D^2*H)", "a*(D^2*H)^b", "a*(D^2*H)^b", "a*D^b*H^c", "a*(D^2*H)^b",
    "a*D^b*H^c", "a*(D^2*H)^b", "a*(D^2*H)^b", "a*(D^2*H)^b", "a*D^b*H^c", "a*D^b*H^c", "a*(D^2*H)^b"),
  a = c(0.000048,0.000064,0.000064,0.000058,0.000059,0.000047,0.000070,
        0.000023,0.000009,-0.147542,0.000007,0.000007,0.000003,0.000010,
        0.000018,0.000031,0.000014,0.000014,0.000011,0.000010,0.000021),
  b = c(1.769901,0.936534,0.936459,0.957300,0.948535,1.677002,0.928364,
        1.016204,1.089575,0.000031,1.121720,1.113125,1.994043,1.074116,
        1.785698,0.987658,1.068129,1.064049,1.960877,1.765752,1.013348),
  c = c(1.165658,NA,NA,NA,NA,1.254950,NA,
        NA,NA,NA,NA,NA,1.644049,NA,
        1.403011,NA,NA,NA,1.308190,1.523197,NA),
  check.names = FALSE
)
knitr::kable(fv, digits = 6)
```

### 2.2 Sharma & Pukkala stem volume

The stem-volume equation is:

$$
V_{s,dm^3}=\exp(a+b\ln D+c\ln H)
$$

followed by the explicit unit conversion:

$$
V_{s,m^3}=\frac{V_{s,dm^3}}{1000}
$$

or equivalently:

$$
V_s=\frac{\exp(a+b\ln D+c\ln H)}{1000}
$$

This `/1000` conversion is part of the implemented calculation and is required because the original Sharma & Pukkala equation reports stem volume in dm$^3$.

The species-specific $a$, $b$, and $c$ coefficients are the same Sharma & Pukkala coefficients listed in the biomass table above.

### 2.3 Forest Regulations 2079 branch volume

Branch volume is calculated from the Sharma & Pukkala stem volume using the branch ratio prescribed in Schedule 9 of Nepal's Forest Regulations 2079:

$$
V_b=R(D)V_s
$$

The DBH-dependent ratio is:

$$
R(D)=s, \quad D<10
$$

$$
R(D)=\frac{(D-10)m+(40-D)s}{30}, \quad 10\le D\le40
$$

$$
R(D)=\frac{(D-40)b+(70-D)m}{30}, \quad 40<D\le70
$$

$$
R(D)=b, \quad D>70
$$

and total tree volume is:

$$
V_t=V_s+V_b
$$

#### Forest Regulations branch groups and parameters

```{r branch-table, echo=FALSE}
br <- forest_regulation_branch_parameters()
names(br) <- c("Branch group", "Species/group", "Nepali name", "s", "m", "b")
knitr::kable(br, digits = 3)
```

The applicable `branch_group` must therefore be supplied for the Sharma–Pukkala + Forest Regulations volume workflow. The package should not infer an unspecified regulatory group from an unsupported species name.

## 3. Plot and forest summaries

For a plot of area $A$ hectares, a tree-level quantity $Y_i$ is expanded as:

$$
Y_{plot}=\frac{\sum_iY_i}{A}
$$

For $n$ plots, the standard error is:

$$
SE=\frac{SD}{\sqrt{n}}
$$

and the two-sided 95% confidence interval for the mean is:

$$
\bar{Y}\pm t_{0.975,n-1}SE
$$

These intervals describe sampling uncertainty in the mean across plots. They are not individual-tree prediction intervals and do not propagate allometric-model uncertainty.

## 4. Method boundaries and interpretation

The methods are not automatically interchangeable estimates of exactly the same biological quantity:

- FRTC biomass excludes the 0–0.30 m stump;
- Chave biomass follows the aboveground-biomass definition of Chave et al. (2014);
- Sharma & Pukkala biomass uses the published air-dry density and component-ratio pathway;
- FRTC volume represents stem volume and excludes branches and the 0.30-m stump;
- FRTC's three volume outputs are distinct definitions;
- Sharma–Pukkala + Forest Regulations total volume includes branch volume; and
- carbon is calculated from biomass rather than independently estimated.

Differences among supported methods can be used to examine sensitivity to methodological choice, but their spread is **not a formal estimate of uncertainty or accuracy**.

## 5. Inspecting methods in R

```{r inspect, eval=FALSE}
frtc_models()
frtc_equation("sal")
frtc_density()
sharma_pukkala_species()
forest_regulation_branch_parameters()
chave_biomass(30, 20, "sal", keep_inputs = TRUE)
allometry_references()
```

## 6. References

The main methodological sources are FRTC (2025), Sharma & Pukkala (1990), Chave et al. (2014), Nepal's Forest Regulations 2079, IPCC (2006), and the Global Wood Density Database v2.2. Run:

```{r references, eval=FALSE}
allometry_references()
```

for the package's complete reference registry and source details.