---
title: "Why nepalallometry?"
description: "Why Nepal needs a transparent and extensible allometry package, and why the first release focuses on biomass."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Why nepalallometry?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why does Nepal need an allometry package?

Forest inventories depend on allometric models to translate field measurements
such as diameter and height into quantities used for management, research, and
carbon accounting. In Nepal, those models are distributed across reports,
regulations, publications, spreadsheets, and separate analytical traditions.
Users must therefore identify the appropriate equation, reproduce its
parameters, apply the correct units and wood density, interpret its calibration
range, and understand exactly which parts of the tree its response represents.

These decisions matter because Nepal's established pathways do not estimate an
identical quantity. They differ in species coverage, moisture basis, density
treatment, calibration information, and biomass boundary. Transferring an
equation without its assumptions can produce a numerical result that appears
precise but is not fully comparable with another estimate. Unsupported species,
unmeasured height, unusual tree form, and application outside the calibration
domain create further practical uncertainty.

`nepalallometry` provides a common, reproducible implementation while preserving
those methodological differences. It records model and density provenance,
identifies unsupported or extrapolated trees, reports model coverage, and keeps
the outputs from different pathways separate. The package is intended to grow
into a broader platform for allometric work in Nepal as additional models and
inventory tools are incorporated.

## First release: biomass estimation

The first release focuses on tree-, plot-, species-, diameter-class-, and
forest-level biomass and carbon estimation. It provides a single inventory
workflow and an auditable Excel workbook designed for researchers, students,
and operational forest managers. Its usefulness for operational decisions will
increase as additional models, species, and allometric modules are added.

The three pathways in this release are intentionally not averaged or combined:

<div class="pathway-grid">
<div class="pathway-card">
<h3>FRTC</h3>
<p>The Forest Research and Training Centre (2025) pathway implements recent species-specific total-biomass equations for seven major tree taxa. The estimates are oven-dry biomass above the 0.30 m felling height, so the lower stump is excluded.</p>
</div>
<div class="pathway-card">
<h3>Sharma &amp; Pukkala</h3>
<p>Sharma and Pukkala (1990) provides air-dry stem, branch, and foliage biomass for named species and explicit miscellaneous groups. Schedule 9 of Nepal's <em>Forest Regulations, 2079</em> provides the regulatory basis for operational use of its tree-volume parameters (Government of Nepal, 2022).</p>
</div>
<div class="pathway-card">
<h3>Chave</h3>
<p>Chave et al. (2014) provides a height-inclusive pantropical oven-dry aboveground-biomass equation. The package supplies traceable basic wood density from the applicable FRTC source or Global Wood Density Database v2.2 matches (Fischer et al., 2026a, 2026b).</p>
</div>
</div>

Carbon is calculated with a default fraction of 0.47 following IPCC (2006).
Users may supply another documented fraction when appropriate. The default is a
reporting conversion and should not be interpreted as a measured
species-specific carbon concentration.

## Who can use it?

- **Researchers** can reproduce calculations and inspect the assumptions,
  calibration flags, and source attached to each pathway.
- **Students** can learn how measurements, equations, density, biomass
  boundaries, and aggregation are connected.
- **Operational managers** can create consistent summaries and retain an audit
  record, while recognizing the present limits of species and model coverage.

## Development perspective

My first research in this area examined forest biomass estimation. I then
continued working on allometric models and found that the practical difficulty
was not simply the absence of equations. Equations already exist, but their
assumptions, response definitions, supporting data, regulatory use, and
appropriate application are often difficult to trace through an operational
workflow. I developed `nepalallometry` to make these choices explicit and to
provide a foundation that can be expanded as new evidence and models become
available.

## References

Complete APA-style references for every model, database, regulatory source, and
conversion factor used by the package are available with:

```{r references, eval=FALSE}
allometry_references()
```

They are also included in `citation("nepalallometry")` and in the `Read_Me`
sheet of each exported workbook.
